---
title: "Exploratory Data Analysis - White Wine Quality"
output: rmarkdown :: github_document
---

```{r Global Options, include=FALSE}
knitr::opts_chunk$set(fig.width = 15,echo=FALSE, warning=FALSE, message=FALSE)
```

```{r Loading necessary libraries}
library(ggplot2)
library(GGally)
library(dplyr)
library(ggcorrplot)
```


```{r Load dataset}
wine_qual <- read.csv('wineQualityWhites.csv')
```

After loading in the dataset, we should start with understanding the composition
, shape of the dataset as well as the datatypes of the different variables

Let us get an idea about the dimensions of the dataset.

```{r Getting dimesions of dataset}
dim(wine_qual)
```


The "quality" variable is obviously a categorical variable and it would be 
beneficial to introduce a new column in the dataset that is a a factor variable 
of the quality with the avaialble levels in the dataset

```{r Create factor quality variable}
wine_qual$quality.factor <- factor(wine_qual$quality)
```

Let us now get an idea of the datatypes of the variables in the dataset after 
adding the new factor variable.

```{r Descriptions of different variables}
str(wine_qual)
```

The chemical properties of the wine are represented as continous variables 
and the quality rating of the wine has been designated by an integer

Quality is the dependent variable being dependent on multiple chemical 
properties of the wine

Let's print out summary of dataset and start concentrating on each variable 
in turn

```{r Summary statistics of different variables}
summary(wine_qual)
```



<h2>Univariate Analysis</h2>

Section Objectives : The objective of this section is to analyse the frequency 
distributions of different variables in the dataset, create coherent plots plots
of these distributions and then make adequate observations about the 
distributions.

Let' start by creating a frequency distribution of the quality.


```{r Histogram of quality factor variable}
hist_quality_factor <- ggplot(aes(x=quality.factor), data=wine_qual) +
  geom_histogram(stat='count')

hist_quality_factor
```

We can see that the distribution has a fairly normal shape with a slight 
right skew.

We can get a better idea of the mean (and this the skew) and median of the 
quality variable (since it is a integral value). For this we need to consider 
the integral "quality variable" in the dataset


```{r Histogram of quality variable}
hist_quality <- ggplot(aes(x=quality), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(breaks=seq(0,max(wine_qual$quality),1)) +
    geom_vline(aes(xintercept=mean(quality),color='mean')) +
  geom_vline(aes(xintercept=median(quality),color='median')) +
  scale_color_manual(name = 'Quality Statistics', values = c(median = 'blue', 
                                                             mean = 'red'))

hist_quality
```


<u>Observation:</u>The distribution shows a somewhat normal trend with most 
wines having been classified as having a quality of 6. We also notice that more 
wines have rating value less than the median rating value.

Next, let's create a freqfrequency distribution of the alcohol content of wines

```{r Histogram of alcohol content variable}
hist_alcohol <- ggplot(aes(x=alcohol), data=wine_qual) +
  geom_histogram(color=I('orange'), fill=NA) +
  scale_x_continuous(breaks=seq(0,max(wine_qual$alcohol),0.5)) +
  geom_vline(aes(xintercept=mean(alcohol),color='mean')) +
  geom_vline(aes(xintercept=median(alcohol),color='median')) +
  scale_color_manual(name = 'Alcohol Statistics', values = c(median = 'blue', 
                                                             mean = 'red'))

hist_alcohol
```

Presented below is the summary of the alcohol content

```{r Summary of alcohol content variable}
summary(wine_qual$alcohol)
```

<u>Observation:</u>The alcohol content distribution is quite interesting. The 
distribution is not quite normal. .We can see two peaks in the distribution, 
one in the 9.0-9.5 alcohol content range and again in the 10.75 to 11.0 interval
.


Let' create a freqfrequency districution of wine pH

```{r Histogram of pH variable}
hist_pH <- ggplot(aes(x=pH), data=wine_qual) +
  geom_histogram(color=I('orange'), fill=NA) +
  scale_x_continuous(breaks=seq(0,max(wine_qual$pH),0.1)) +
  geom_vline(aes(xintercept=mean(pH),color='mean')) +
  geom_vline(aes(xintercept=median(pH),color='median')) +
  scale_color_manual(name = 'pH Statistics', values = c(median = 'blue', 
                                                        mean = 'red'))

hist_pH

```

Presented below is the summary of the pH values

```{r Summary of pH variable}
summary(wine_qual$pH)
```

<u>Observation:</u> The wine pH variable has a substantial normal distribution 
shape. There are also no discernable tails in the distribution. The pH level 
varies from 2.72 to 3.82 and this tallies with some additional research that I
did which suggests that most wines have a pH value between 2.8-3.9

Let us now analyse the frequency distribution of the residual sugar amounts in 
the wines.

```{r Histogram of residual sugar variable}

hist_residual_sugar <- ggplot(aes(x=residual.sugar), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(breaks=seq(0,max(wine_qual$residual.sugar),1),
                     limits=c(0,quantile(wine_qual$residual.sugar,0.99))) +
  geom_vline(aes(xintercept=mean(residual.sugar),color='mean')) +
  geom_vline(aes(xintercept=median(residual.sugar),color='median')) +
  scale_color_manual(name = 'Residual Sugar Statistics', values = c(
    median = 'blue', mean = 'red'))

hist_residual_sugar
```


Presented below is the summary of the residual sugar content

```{r Summary of residual sugar variable}
summary(wine_qual$residual.sugar)
```

<u>Observation:</u>The distribution for the residual sugar amount is a prime 
example of a long-tailed distribution. A majority of the wine samples have sugar
contents in the 1-3 range.

Let us log transform (10 base) the distribution to get a better, garnular idea 
of potential peaks

```{r Histogram of residual sugar variable log transformed}

hist_residual_sugar_log <- ggplot(aes(x=residual.sugar), data=wine_qual) +
  geom_histogram() +
  scale_x_log10(breaks=seq(0,max(wine_qual$residual.sugar),1),
                limits=c(1,quantile(wine_qual$residual.sugar,0.95))) +
  geom_vline(aes(xintercept=mean(residual.sugar),color='mean')) +
  geom_vline(aes(xintercept=median(residual.sugar),color='median')) +
  scale_color_manual(name = 'Residual Sugar Statistics', values = c(
    median = 'blue', mean = 'red')) +
  labs(x='residual sugar (log_10)')

hist_residual_sugar_log
```

<u>Observation:</u>

Let' create a freqfrequency districution of free and total suphur dioxide

```{r Histogram of sulphur dioxide contents}

library(gridExtra)

plot_total_so2 <- ggplot(aes(x=total.sulfur.dioxide), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(breaks=seq(0,max(wine_qual$total.sulfur.dioxide),50),
                     limits = c(0,quantile(wine_qual$total.sulfur.dioxide,0.99)
                                )) +
  geom_vline(aes(xintercept=mean(total.sulfur.dioxide),color='mean')) +
  geom_vline(aes(xintercept=median(total.sulfur.dioxide),color='median')) +
  scale_color_manual(name = 'Total SO2 Statistics', values = c(
    median = 'blue', mean = 'red'))


plot_free_so2 <- ggplot(aes(x=free.sulfur.dioxide), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(breaks=seq(0,max(wine_qual$free.sulfur.dioxide),25),
                limits = c(0,quantile(wine_qual$free.sulfur.dioxide,0.99))) +
  geom_vline(aes(xintercept=mean(free.sulfur.dioxide),color='mean')) +
  geom_vline(aes(xintercept=median(free.sulfur.dioxide),color='median')) +
  scale_color_manual(name = 'Free SO2 Statistics', 
                     values = c(median = 'blue', mean = 'red'))

label_plot <-   scale_color_manual(name = 'Free SO2 Statistics', values = c(
  median = 'blue', mean = 'red'))

grid.arrange(plot_total_so2,plot_free_so2, ncol=2)

```



Presented below is the summary of the total sulfur dioxide content

```{r Summary of total sulphur dioxide contents}
summary(wine_qual$total.sulfur.dioxide)
```

and the free sulfur dioxide content

```{r Summary of free dioxide contents}
summary(wine_qual$free.sulfur.dioxide)
```

<u>Observation:</u> While the total sulfur dioxide distribution is quite 
straightforward with no noticeable skew, the free sulfur dioxide distribution 
seems to almost bimodal with two distinct peaks at the 20 and 35 free sulfur 
dioxide levels. There is also a sharp and noticeable dip at the 32-34 level of 
the distribution.

Let us now create the distribution for chlorides in the wine samples in the 
dataset

```{r Histogram of chlorides variable}
hist_chlorides <- ggplot(aes(x=chlorides), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(
    limits=c(0.0,quantile(wine_qual$chlorides,0.99))) +
  geom_vline(aes(xintercept=mean(chlorides),color='mean')) +
  geom_vline(aes(xintercept=median(chlorides),color='median')) +
  scale_color_manual(name = 'Chlorides Statistics', values = c(
    median = 'blue', mean = 'red'))

hist_chlorides
```


Presented below is the summary of the chlorides

```{r Summary of chlorides variable}
summary(wine_qual$chlorides)
```

<u>Observation:</u>We can see that majority of wines have a chloride level less 
than 0.10. The distribution is also a long tailed one with a distinctive right 
skew.

Another final variable of interest is the density. Let us also look into the 
frequency distribution for the density.

```{r Histogram of density variable}
hist_density <- ggplot(aes(x=density), data=wine_qual) +
  geom_histogram() +
  scale_x_continuous(
    limits=c(0.99,quantile(wine_qual$density,0.99))) +
  geom_vline(aes(xintercept=mean(density),color='mean')) +
  geom_vline(aes(xintercept=median(density),color='median'), linetype=2) +
  scale_color_manual(name = 'Density Statistics', values = c(median = 'blue', 
                                                             mean = 'red'))

hist_density
```


Presented below is the summary of the density of the wine

```{r Summary of density variable}
summary(wine_qual$density)
```

<u>Observation:</u>The density distribution shows a distinct decrease in count 
with increasing density although the decline is gradual rather than steep (such 
as one that creates a long tailed distribution)

This concludes our Univariate Analysis section. Presented below are some of the 
relevant questions asked and the answers thereof

<b>What are the variables of most interest in the dataset?</b> 
</br>
Out of the variables considered in the section above, I suspect that the alcohol
level of the wines, pH level of the wines, residual sugar content and density of
the wines will be the most important determinants of the wine's quality. Further
analysis in the following sections will either confirm of deny my suspicions. 
Other variables that may have a lesser impact are the chloride level and the 
level of free and total sulfur dioxides in the wine.

<b>Were any new/calculated variables created?</b>
</br>
For the purposes of the analysis I have created a new factor variable of the 
quality out of the continous variable of the same present in the dataset. This 
factor variable will be of interest in the Bivariate analysis section as well.

<b>Were there any unusual distributions? Were any transformations made to 
accommodate for such data?</b>
</br>
While creating histograms for the different continuous variables, I noticed 
that the residual sugar distribution was distinctly long tailed. Therefore I log
transformed the distribution to get a better idea of the distribution. 
As a result, I was able to discern three distinctive peaks in the 
frequency distribution. One at the 1.5 residual sugar level, one at 7.5 and the
last one at the 13.5 residual sugar level.

<h2>Bivariate Analysis</h2>

Section Objectives : The objective of this section is to analyse the 
co-relations between pairs of variables in the dataset. The pairs of variables 
may or may not include the dependent variable vector viz. the wine quality.

The section will also include analysis of the trends between variables and 
analysis based on conditional means and medians. This will give us a more 
hollistic idea of the trends between variables and allow us to tackle issues 
such as the bias/variance tradeoff.

A quick and convenient way to ascertain the co-relations between variables is to
create a corelation matrix by which we can focus in on the individual 
relationships between the continuous variables. This also makes sense as within 
the dataset are continous variables and there are no categorical/factor 
variables present so we can expect a numerical co-relation value between pairs 
of variables.


```{r Co-relation matrix of different variables}
data_cor <- wine_qual %>% select(-quality.factor)
corr <- round(cor(data_cor),1)
ggcorrplot(corr, hc.order=TRUE, type='lower', lab=TRUE)
```


As we can see from the matrix produced there does not seem to be a very high 
level of co-relation between the other variables and the quality of the white 
wine sample. The strongest co-relation(which by itself is quite a moderate one) 
is between the alcohol content of the wine and its quality. This can providde 
us with a starting point for our investigations.

<h3>Analysis of different variable pairs</h3>

Lets begin the analysis by creating boxplots of all of the wine qualities 
against each of the independent variables. We will use our factor variable of 
qualities that we created in the dataset earlier.

```{r Boxplots of different variables across quality values}
var_names <- data.frame(names(wine_qual))

var_names_fil <- var_names %>%
  filter(!names.wine_qual. %in% c('X','quality','quality.factor'))

var_names_fil <- var_names_fil$names.wine_qual.

p=list()

for (i in var_names_fil){
  p[[i]] <- ggplot(aes_string(x='quality.factor', y=i, color='quality.factor'), 
                   data=wine_qual) +
    geom_boxplot(show.legend = FALSE)

}

do.call(grid.arrange,c(p, padding=5))
```


From the above plots we can see that for most of the variables the variations 
in central tendency(in this case the median) is not pronounced. However we can 
find a distinct trend of increasing quality with increasing alcohol content and
decreasing quality with increasing density. Let's plot out the co-relation 
between these variables and more in greater detail.


Let's explore these two in detail.

We start with the boxplot for alcohol content vs quality

```{r Boxplot of alcohol variable with quality factor variable}
ggplot(aes(x=quality.factor, y=alcohol, color=quality.factor), data=wine_qual) +
  geom_boxplot() +
  stat_summary(fun.y=median, geom='line', aes(group=1)) +
  stat_summary(fun.y=median, geom="point")
```

We can see that after the quality factor of 5, for all higher qualities, the 
medians of the alcohol contents increase consistently with quality. We can also
make an additional observation i.e quality factor 5 has significantly higher 
outliers than the other quality factor groups.

Here is a representation of the boxplot data in tabular form.


```{r Summary of alcohol variable across quality levels}
wine_qual %>%
  group_by(quality.factor) %>%
  summarise(min=min(alcohol),
            first_quantile=quantile(alcohol,0.25),
            median=median(alcohol),
            third_quartile=quantile(alcohol,0.75),
            max=max(alcohol))
```



Let's consider the density boxplot now

```{r Boxplot of density variable with quality factor variable}
ggplot(aes(x=quality.factor, y=density, color=quality.factor), data=wine_qual) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0.99,1.01)) +
  stat_summary(fun.y=median, geom='line', aes(group=1)) +
  stat_summary(fun.y=median, geom="point")
```


In this plot, we see a pattern of consistently decreasing median of the density 
with increasing wine quality.

A tabular representation of the density boxplot data has been presented below.

```{r Summary of density variable across quality levels}
wine_qual %>%
  group_by(quality.factor) %>%
  summarise(min=min(density),
            first_quantile=quantile(density,0.25),
            median=median(density),
            third_quartile=quantile(density,0.75),
            max=max(density))
```

Now let us start comparing individual datapoints in the dataset and discern 
possible co-relations in between pairs of variables.

Let us begin with the quality vs alcohol content variable pair. Provided below 
is a plot (jitter being applied to deal with overplotting) between alcohol 
content and quality of the wine.

```{r Alcohol content vs quality}
ggplot(aes(x=alcohol, y=quality), data=wine_qual) + 
  geom_jitter(alpha=0.05) +
  scale_x_continuous(breaks=seq(0,max(wine_qual$alcohol),1)) +
  scale_y_continuous(breaks=seq(0,max(wine_qual$quality),1)) +
  stat_smooth(method='lm')
```


By itself, the above plot is not too insightful. The linear trend line shows us 
a feeble positive co-relation which doesn't say anything we don't already know.

Maybe we could aggreagate over the alcohol content to obtain conditional mean 
and median qualities. A similar trend would only reaffirm our suspicions.

Before we do so we need to perform an analysis of the unique alcohol values so 
as to not obtain a graph with high degrees of variance.

Lets find out the number unique values of alcohol content and their distribution

```{r Statistics of alcohol variable}
length(unique(wine_qual$alcohol))  

table(wine_qual$alcohol)
```


With a 103 unique values the plot should not lend itself to high degrees
of overplotting. Let's use the dplyr package to aggregate over the alcohol 
content and find mean and median quality values for each inviduidual alcohol 
content value.

```{r Conditional means and variance of Quality with alcohol content}
library(dplyr)
wine_qual.grp_by_alcohol <- wine_qual %>%
  group_by(alcohol) %>%
  summarise(mean_quality=mean(quality),
            median_quality=median(quality),
            variance_quality=var(quality),
            n=n()) %>%
  arrange(alcohol)
```

Now let us create the plot of the conditional means of the wine quality.

```{r Plot of conditional means of Quality with alcohol content}
ggplot(aes(x=alcohol, y=mean_quality), data=wine_qual.grp_by_alcohol) +
  geom_line() +
  geom_smooth()
```

Let us also perform a co-relation test between the alcohol content groups 
and the graph for the conditional means of quality
```{r Co-relation co-efficient of conditional means of quality with alcohol}
cor.test(x=wine_qual.grp_by_alcohol$alcohol, 
         y=wine_qual.grp_by_alcohol$mean_quality)
```

The co-relation co-efficient shows us a greater degree of co-relation than for 
the individual alcohol vs quality plot.

We can also plot for the conditional variance. But we have to be careful here. 
For alcohol contents with only one observation the variance is calculated as NA.
We should disregard these values before devising our plot. Even after the 
exclusion of these values we can have a distinct trend of the conditional 
variance of quality with respect to alcoohol content.

```{r Plot of conditional variance of Quality with alcohol content}
ggplot(aes(x=alcohol, y=variance_quality), 
       data=subset(wine_qual.grp_by_alcohol, !is.na(variance_quality))) +
  geom_line()+
  geom_smooth()
```


From the plot above we can deduce that wines with high alcohol contents have 
relatively lower variations among their quality values. The disparity is 
especially visible in wines with alcohol content less that 10 and thereafter 
the dispersion is not as pronounced.

A similar co-relation that may be explored is between wine density and quality. 
Let us begin the same way we did with alcohol vs quality and create a basic 
plot. Let us add jitter and opacity to deal with overplotting.

```{r Density vs quality}
ggplot(aes(x=density, y=quality), data=wine_qual) + 
  geom_point(position='jitter',alpha=0.05) +
  scale_x_continuous(breaks=seq(0,max(wine_qual$density),0.005),
                     limits=c(0.985,1.005)) +
  scale_y_continuous(breaks=seq(0,max(wine_qual$quality),1)) +
  stat_smooth(method='lm')
```


Another pair of variables that would be interesting for analysis is the alcohol 
vs density. These two variables have a strong negetive co-relation between them
and density has the second best co-relation with quality among the other 
variables. It would be interesting to see the and analyse the co-relation 
between these two variables. Later, as part of our multivariate analysis we can
consider including alcohol, denisty and quality and observe their 
inter-relationships.



```{r Plot of Alcohol content vs density}
ggplot(aes(x=alcohol, y=density), data=wine_qual) +
  geom_point()
```

The plot definitely suffers from overplotting. Let us rectify that by adding 
jitter and opacity to the datapoints. Also, since we have very few values of 
density greater than 1.01, a few adjustments to the y-axis will create a more 
robust plot.

```{r Refined plot of Alcohol content vs density}
ggplot(aes(x=alcohol, y=density), data=wine_qual) +
  geom_point(position='jitter', alpha=0.05) +
  scale_y_continuous(breaks=seq(0.99,1.01,0.005),
                     limits=c(0.99,1.01)) +
  geom_line(stat='summary', fun.y=mean, color='red') +
  geom_smooth(size=2)
```


From the plot we can distinctively notice the negative co-relation between the 
density and alcohol content of the wines. The plot further shows us that 
majority of the datapoints are confined with alcohol contents less than roughly 
11.5.


Anoher feature that has a high co-relation co-efficient(in magnitude) with the 
density is the residual sugar content. This co-relation can also be the basis of
any further co-relation between residual sugar and alcohol content. Let us first
look at density and residual sugar.


```{r Plot of Density vs residual sugar content}
ggplot(aes(x=density, y=residual.sugar), data=wine_qual) +
  geom_point()
```


Just as in the plot for density and alcohol, changes for accomodating 
overplotting and axis modifications need to be made for creating a clearer plot.

```{r Refined plot of Density vs residual sugar content}
ggplot(aes(x=density, y=residual.sugar), data=wine_qual) +
  geom_point(position='jitter', alpha=0.1, color='orange') +
  scale_x_continuous(limits=c(0.99,1.005)) +
  scale_y_continuous(limits=c(0,25)) +
  geom_smooth(span=10) +
  geom_rect(xmin=0.990, xmax=0.995, ymin=0, ymax=2.5, fill=NA, color='black')
```


The plot and the smoothened curve gives us the high positive co-relation we were
expecting. There is however one interesting observation. There is a significant
clutter of datapoints with density less than 0.995 (highlighted in plot) that do
not have a strong positive co-relation. These points seem to maintain a constant
residual sugar value with increasing density. It would be interesting to analyse
the cause of the behaviour of these points and whether any other variables are 
responsible for this effect. We can take up this analysis in the multivariate 
analysis section.


A final analysis is between the variables free sulfur dioxide and total sulfur 
dioxide. The variables obviously have a positive co-relation since total sulfur 
dioxide will obviously increase with increased free sulfur dioxide. 

Lets create one such plot and discuss its implications.

```{r Total sulfur dioxide vs Free sulfur dioxide}
ggplot(aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide), data=wine_qual) +
  geom_line(color='yellow',position=position_jitter(w=0.01, h=0)) +
  stat_smooth(method='lm', span=20)
```

The co-relation is unmistakable and it should be so. In fact after a 
free_sulfur_dioxide content of ~140, we see an almost linear trend. However, 
this co-relation may be of consequence when trying to create predictive models 
based on multiple linear regression methods. The level of multicollinearity 
between these two variables may create some level of unpredictability in the 
models and therefore it would be best to avoid one of these variables when 
creating such models.

Another such inferable relationship should be between the different acidities 
and the pH value. A higher acidity does suggest a lower pH value and we 
actually see such a trend for all of the acidities and pH level. The strongest
co-relations seems to be between the fixed acidity level and the pH. Let us 
consider this relationship through a plot

```{r Fixed acidity vs pH}
ggplot(aes(x=pH, y=fixed.acidity), data=wine_qual) +
  geom_point(position='jitter',alpha=0.1, color='orange')+
  geom_smooth(color='blue', linetype=2)
```

As expected we do notice a negative co-relation. We must be careful to 
disregard the region created by the smoothened trned line outside the data 
point space. Even so we do notice our speculated co-relation.

This concludes our Bivariate Analysis section. Presented below are some of the
relevant questions asked and the answers thereof

<b>What are the co-relations/relationships that are observed between different
pairs of variables in the dataset?</b> 
</br>
There are quite a few analyses that have been performed in the bivariate 
analysis section. They are listed below
<ol>
<li>With respect to the dependent variable viz. the wine quality, the variable
that has the most discernable co-relation is the alcohol content of the wine. 
The boxplot reveals that with increasing wine quality from rating 5, the median
of the alcohol content also increases. Below quality level 5 however, 
the co-relation is a negative one. But for most wines, the quality of the wine
is directly co-related with the alcohol content.</li>
<li>We also see that with increasing wine quality, the density of the wine 
decreases. This co-relation affirms that better wines are more fluid.</li>
<li>The co-relation between alcohol content and density is also quite strong 
and negative. More alcoholic wines are less dense throughout the dataset</li>
<li>The distribution between density and residual sugar is also quite marked. 
The c0-relations shows that sweeter wines have a higher density. There is a 
segment of wines with lower densities have more or less constant residual 
sugar amounts with increasing density. We will take up this co-relation in the
multivariate analysis section</li>
<li>We also explored the co-relation between the free and total sulfur dioxide
contents and found out that, as expected, total sulfur dioxide increases with
free sulfur dioxide content, indicating a strong positive co-relation. </li>
</ol>

<b>What variable has the strongest impact on the dependent variable? What are 
some other strong relationships that exclude the dependent variable</b>
</br>
From the analysis above, we can discern that the alcohol level has the 
strongest co-relation with the wine quality level.

Other strong co-relations included the ones between alcohol content and 
density, between residual sugar content and density, between total sulfur 
dioxide and free sulfur dioxide and between fixed acidity and pH level of wines.


<h2>Multivariate Analysis</h2>


Section Objectives : In this section we intend to advance our investigation 
into the variables and consider the relationships thath exist between more than
two variables at a time. This will give us a better understanding of the ways
in which the variables relate to each other and influence each other.

To begin with we can take up the triplet of quality, alcohol content and 
density. Let us create a scatterplot and observe the co-relation between the
three. Also we will include points whose density values are less than the 
99.9th percentile for density values. That way, we would certainly not be 
excluding any significant number of points and would also be able to create a
discernable contrast of colors to highlight the density variation.

```{r  Alcohol contents vs quality vs density}
library(viridis)
ggplot(aes(x=alcohol, y=quality), 
       data=subset(wine_qual,density<=quantile(density,0.999))) +
  geom_point(aes(colour=density), position='jitter') +
  geom_smooth(color='red') +
  scale_color_viridis(discrete=FALSE, option='magma')
```




From the trend line in this plot we can see that wines with greater alcohol 
content generally have higher quality rating(as seen from the trend line) 
whereas they have lower density.

Further let us perform the same analysis for the quality, alcohol and residual
sugar amounts. We can expect to see a similar trend. We will also consider
values of residual sugar with values less than the 99.9th percentile for 
reasons stated in the density plot above.

```{r Alcohol contents vs quality vs residual sugar}
ggplot(aes(x=alcohol, y=quality, color=residual.sugar), data=subset(wine_qual, residual.sugar<=quantile(residual.sugar,0.999))) +
  geom_point(position='jitter') +
  geom_smooth(color='blue') +
  scale_color_viridis(discrete=FALSE, option='plasma')
```


Inaddition to the already known positive co-relation between alcohol content
and quality we also see that that wines with greater alcohol content have 
lesser levels of residual sugar and can thus be assume to be less sweet.

Let us now consider the three variables of alcohol, density and residual sugar.

```{r Alcohol contents vs residual sugar vs density}
ggplot(aes(x=alcohol, y=density, color=residual.sugar), 
       data=subset(wine_qual, residual.sugar<=quantile(residual.sugar,0.999))) +
  geom_point(position='jitter') +
  scale_y_continuous(limits=c(0.99,1.01)) +
  geom_smooth(color='blue', alpha=0.05) +
  scale_color_gradientn(colours = heat.colors(10))
```

The plot cleary shows what we already know. Sweeter wines have higher densities
but lower alcohol contents.

We should also check to see if this co-relation between density and residual
sugar holds for each individual quality level.

```{r Density vs quality vs density}
ggplot(aes(x=quality.factor, y=density, color=residual.sugar), data=wine_qual) +
  geom_point(position='jitter') +
  scale_y_continuous(limits=c(0.99,1.01)) +
  geom_smooth(color='blue', alpha=0.05) +
  scale_color_gradientn(colours = rainbow(10))
```

Except for the quality levels of 3 and 9 (where there are too few datapoints
to make concrete conclusions), we can see that indeed with increasing density
residual sugar levels in wines do tend to increase.

It would be good to have a greater level of granularity for this plot for each
quality level.
```{r Alcohol content vs density vs residual sugar across different quality}
ggplot(aes(x=alcohol, y=density, color=residual.sugar), data=subset(wine_qual, residual.sugar<=quantile(residual.sugar,0.999))) +
  geom_point(position='jitter') +
  geom_smooth(color='red') +
  facet_wrap(~quality.factor) +
  scale_color_viridis(discrete=FALSE, option='magma')
```

The quality levels of 3 and 9 hardly have enough data points for consideration.
For the other quality levels we observe see that the number of wines with 
higher residual sugar contents decreases with increasing quality. We also 
observer that for a particular quality level, residual sugar level decreases
with increasing alcohol content. These observations supplement the infomation 
we have obtained from the previous plot.

We finish this section with an analysis of residual sugar vs density that we
took up in the bivariate analysis section. We saw that for lower densities 
residual sugar did not have a significant positive co-relation with density.
To figure out the cause we should look into two variables in particular viz the
total_sulfur_dioxide and the alcohol content in wines. We plot the graphs in 
question.

```{r Density vs residual sugar with alcohol and total sulfur dioxide}
res_sugar_density_so2 <- ggplot(aes(x=density, y=residual.sugar, 
                                color=total.sulfur.dioxide), data=wine_qual) +
  geom_point(position='jitter', alpha=0.1) +
  scale_x_continuous(limits=c(0.99,1.005)) +
  scale_y_continuous(limits=c(0,25)) +
  geom_smooth(span=10) +
  scale_color_gradientn(colors=terrain.colors(100)) +
  geom_rect(xmin=0.990, xmax=0.995, ymin=0, ymax=2.5, fill=NA, color='black')

res_sugar_density_alcohol <- ggplot(aes(x=density, y=residual.sugar, 
                                        color=alcohol), data=wine_qual) +
  geom_point(position='jitter', alpha=0.1) +
  scale_x_continuous(limits=c(0.99,1.005)) +
  scale_y_continuous(limits=c(0,25)) +
  geom_smooth(span=10) +
  scale_color_gradientn(colors=heat.colors(100)) +
  geom_rect(xmin=0.990, xmax=0.995, ymin=0, ymax=2.5, fill=NA, color='black')

grid.arrange(res_sugar_density_so2,res_sugar_density_alcohol, nrow=2)
```



Now both alcohol and total sulfur dioxide have negative co-relations with 
residual sugar quantity for wines with lower density. It is thus possible 
that a combination of both higher alcohol volume and lower total sulfur dioxide
levels for wines with lower densities gives rise to the lessned residual sugar
amounts. Although it would be incorrect to assume that any one variable is more
significant than the other.

This concludes our Multivariate Analysis section. Presented below is the 
summary of this section.

<b>What are the co-relations/relationships that are observed between different
variables in this multivariate analysis section?</b> 
</br>
In this section we have dealt with the relationships among multiple variables.
Observed co-relations are as below
<ol>
<li>We see that quality of wines, intially drops a tad with increasing alcohol
content but for the majority of wines in the dataset, quality increases with 
increasing alcohol content. At the same time we see a distinct drop in density
for wines that have more alcohol.</li>
<li>We also see a similar pattern for the residual sugar content of wines. With
increasing alcohol content the sweetness of wines falls.</li>
<li>From the third and fourth plots, we are able to obtain certain additional
information in addition to what has been detailed above. It is that for a 
particular alcohol content the residual sugar content tends to increase with
density of the wine. The same is also true for a particular quality level of 
wine.</li>
<li>We can also discern a co-relation between total sulfur dioxide content and
density. we see that sulfur dioxide contents of wines tend to increase with 
density.</li>
<li>An important onservation is that for a certain region of ower density wines
, residual sugar content tends to remain constant with increasing density. This
<b>may be</b> due to the effect of lower sulfur dioxide content and higher 
alcohol contents in these wines but we really cannot conclude any concrete 
causation.</li>
</ol>


<h2>Final Plots and Summary<h2>

<h3>Plot One</h3>

```{r Histogram of residual sugar variable log transformed final}

ggplot(aes(x=residual.sugar), data=wine_qual) +
  geom_histogram() +
  scale_x_log10(breaks=seq(0,max(wine_qual$residual.sugar),1),
                limits=c(1,quantile(wine_qual$residual.sugar,0.95))) +
  labs(x='Residual Sugar Amount', y='Number of Wines', 
       title='Log_10 Residual Sugar')
```

<h3>Description  for Plot One</h3>

The right-skewed residual sugar histogram has been log transformed(base 10) to
get a better idea of the frequency. Thus we caan spot three distonctive peaks 
in the distribution at residual sugar levels of 1.5 7.5 and 13.5.


<h3>Plot Two</h3>

```{r Alcohol content vs quality final}

bp1 <- ggplot(aes(x=quality.factor, y=alcohol, color=quality.factor), 
              data=wine_qual) +
  geom_boxplot() +
  stat_summary(fun.y=median, geom='line', aes(group=1)) +
  stat_summary(fun.y=median, geom="point") +
  labs(x='Quality', 
       y='Alcohol Content', 
       title='Alcohol Content distribution by quality level')


bp2 <- ggplot(aes(x=alcohol, y=mean_quality), data=wine_qual.grp_by_alcohol) +
  geom_line() +
  geom_smooth() +
  labs(x='Alcohol Content', y='Mean of Quality level', 
       title='Conditional Mean of wine quality by alcohol level')

grid.arrange(bp1, bp2, ncol=1)
```

<h3>Description  for Plot Two</h3>

The plots above work in conjunnction to display the fact that increasing 
alcohol content is linked to higher wine quality. The first boxplot shows the 
increasing trend of median alcohol content with increasing wine quality. The
second plot also bears testimony to this relationship. With increasing alcohol
levels the conditional means of quality levels also increases.

<h3>Plot Three</h3>

```{r Alcohol contents vs residual sugar vs density final}
ggplot(aes(x=alcohol, y=density, color=residual.sugar), 
       data=subset(wine_qual, residual.sugar<=quantile(residual.sugar,0.999))) +
  geom_point(position='jitter') +
  scale_y_continuous(limits=c(0.99,1.01)) +
  geom_smooth(color='blue', alpha=0.05) +
  scale_color_gradientn(colours = heat.colors(10), 
                        name='Residual Sugar Content') +
  labs(x='Alcohol Content', y='Density', 
       title='Density vs Alcohol Content and Residual Sugar')
```

<h3>Description  for Plot Three</h3>

The plot reveals that wines with a higher alcohol content typically have lower 
densities. The plot also shows that with increasing density the sweetness of 
wines (represented by residual sugar amount) also increases. We can summarily 
deduce that for a greater alcohol content a greater proportion of wine samples 
will be less sweet and 
will be less dense.   

<h2>Reflection</h2>

The white wine dataset contains 4898 observations on different chemical 
properties of white wine that act as a factor to its overall quality. 
A noticeable feature of the dataset is the absence of any categorical 
variables.I started the exploratory data analysis by considering individual 
variables and their frequency distributions and then took the analysis further
by considering the relationships between pairs of and even more than two of 
these variables and touched on interesting questions pertaining to the data.

On detailed analysis of the dataset, I came to realise that alcohol content 
plays the most crucial role in determing wine quality. The alcohol content also
strongly co-relates with the density of the wine and a higher alcohol content
usually means a less dense wine. The specific gravity of ethanol(0.791) is less
than that of water and it makes sense that a larger alcohol (ethanol) content 
would reduce the overall density of the wine. Likewise an increase in the sugar
content of the wine also adds to its density and thus the two are positively 
co-related. This co-relation may also be influenced by the level of total 
sulfur dioxide and alcohol in the wine and such an effect has been addressed
in the appropriate section of the multivariate analysis. Also I noticed 
certain common sense relationshis such as the one between total sulfur dioxide
and free sulfur dioxide as well as between fixed acidity levels and pH levels.

The analysis is not without its limitations. All relationships among variables
are only co-relations and no causative effects can be inferred for any 
particular trend in any of the variables and their relationships.Additionally
I would have liked to analyse the effect of non checmical properties such as 
vineyard and period of fermentation on the wine quality as they may have 
significant effects on the quality. I would really like, in a subsequent 
analysis, to acquire and consider this data for a more rounded analysis.

